---
title: "Forecasting for Economics and Business"
subtitle: "Lecture 2: Time Series"
author: "David Ubilava"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, style.css]
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 11, fig.height = 7)
```


```{r echo=FALSE, include=FALSE, message=FALSE}
library(ggplot2)
```


# Time Series

Time series are realizations of a chronologically stored sequence of random variables. This sequence of random variables is referred to as the *stochastic process*. Thus, a time series is a realisation of the stochastic process. 

We index time periods as $1,2,\ldots,T$, and denote the set of observations as $\{y_t: t=1,\ldots,T\}$. A time series is a finite sample from an underlying doubly--infinite sequence: $\{\ldots,y_{-1},y_{0},y_1,\ldots,y_{T-1},y_{T},y_{T+1},\ldots\}$.

---


# White Noise

The simplest kind of time series is comprised of realisations from an independent and indentically distributed random variable with zero mean and constant variance: $\varepsilon_t \sim iid\left(0,\sigma^2\right)$. 

The next graph plots this series against time.

---


# White Noise

```{r echo=FALSE, message=FALSE}
n <- 120

set.seed(n)
y <- rnorm(n)

df <- data.frame(t=c(1:n),y=y)

ggplot(df,aes(x=t,y=y))+
  geom_line(size=1,col="skyblue")+
  labs(y=expression(paste(y[t],sep="")))+
  theme_classic()+
  theme(axis.title = element_text(size=18),axis.text = element_text(size=14))
```

---


# White Noise

Such time series are referred to as *white noise*. That is, a time series, $y$, is a white noise process if: 
\begin{align*}
& E(y_t) = 0,\;~\forall~t\\
& Var(y_t) = \sigma^2,\;~\forall~t\\
& Cov(y_t,y_{t-k}) = 0,\;~\forall~k \ne 0
\end{align*}

Because each observation is drawn from the same distribution, white noise is a *stationary* time series. Indeed, it is a special type of stationary time series insofar as it has time-invariant mean, variance, and covariance. 

Stationarity is an important feature, and the assumption, on which the time series analysis heavily relies. 

---


# Stationarity

For a covariance stationary (or, weakly stationary) time series, the mean doesn't need to be equal to zero, it only needs to be constant over time; nor covariances need to be equal to zero, they also only need to be constant over time, though they may vary with $k$. 

Thus, a time series, $\{y_t\}$ is stationary if the mean and variance are independent of $t$, and the autocovariances are independent of $t$ for all $k$.

---


# Autocorrelation

It is more of a norm, rather than an exception, for a time series to be correlated. Indeed, we often observe dependence among the temporally adjacent time series. That is, for reasonably small values of $k$, $y_t$ and $y_{t-k}$ tend to be correlated. Such correlations are referred to as *autocorrelations*, and are given by: $$\rho_k \equiv Corr(y_t,y_{t-k}) = \frac{Cov(y_t,y_{t-k})}{Var(y_t)},\;~~k=1,2,\ldots$$ 

---


# Autocorrelation

We can illustrate the time series autocorrelations using *autocorrelogram*, which plots the sequence of autocorrelation coefficients against the lags at which these coefficients are obtained. 

Under the null of independence, the estimated autocorrelations, $\hat{\rho}_k$, are asymptotically standard normally distributed, $\sqrt{T}\hat{\rho}_k\xrightarrow{d}N(0,1)$, and thus: $\hat{\rho}_k \sim N(0,T^{-1})$

The approximate $(1-\alpha)$% confidence interval is bounded by $-z_{\alpha/2}T^{-1/2}\;\text{and}\;z_{\alpha/2}T^{-1/2}$

The next graph plots an autocorrelogram of the previously illustrated white noise process along with the 95% confidence interval.

---


# Autocorrelation

```{r echo=FALSE, message=FALSE}
maxlag <- 10
df <- data.frame(k=c(1:maxlag),rho=c(acf(y,plot=F)[1:10]$acf))

ggplot(df,aes(x=k,y=rho))+
  geom_segment(aes(xend=k,yend=0))+
  geom_hline(yintercept=0,size=.8)+
  geom_hline(yintercept=c(-1.96/sqrt(n),1.96/sqrt(n)),size=.5,linetype=5,col="skyblue")+
  scale_x_continuous(breaks=c(1:10),labels=c(1:10))+
  labs(x="k",y=expression(rho[k]))+
  coord_cartesian(ylim=c(-1,1))+
  theme_classic()+
  theme(axis.title = element_text(size=18),axis.text = element_text(size=14))
```

---


# Transformations

Sometimes we (have to) transform time series, either by first-differencing or by taking natural logarithms (or both), to work with a suitable variable for the desired econometric analysis. 

The first-difference operator is denoted by $\Delta$, so that $\Delta y_t = y_t-y_{t-1}$. First-differencing helps remove a linear trend from the time series.

If an economic time series is characterized by an apparent exponential growth (e.g., real GDP), by taking natural logarithms the time series 'flatten' and the fluctuations become proportionate. 

The next three graphs illustrate (i) a time series with an apparent exponential growth, (ii) the natural logarithms of this time series, and (iii) the first-differences of the log--transformed series.

---


```{r echo=FALSE, message=FALSE}
n <- 120
a <- 0.05
b <- 0.05
s <- 0.05
set.seed(n)
e <- rnorm(n,0,s)
dlny <- e
for(i in 2:n){
  dlny[i] <- .05+.05*dlny[i-1]+e[i]
}
lny <- cumsum(dlny)
y <- exp(lny)

df <- data.frame(t=c(1:n),y=y,lny=lny,dlny=dlny)
```

# Transformations

```{r echo=FALSE, message=FALSE}
ggplot(df,aes(x=t,y=y))+
  geom_line(size=1,color="skyblue")+
  labs(y=expression(paste(y[t],sep="")))+
  theme_classic()+
  theme(axis.title = element_text(size=18),axis.text = element_text(size=14),axis.text.y = element_text(angle=90,hjust=0.5))
```

---


# Transformations

```{r echo=FALSE, message=FALSE}
ggplot(df,aes(x=t,y=lny))+
  geom_line(size=1,color="skyblue")+
  labs(y=expression(paste("ln",y[t],sep="")))+
  theme_classic()+
  theme(axis.title = element_text(size=18),axis.text = element_text(size=14),axis.text.y = element_text(angle=90,hjust=0.5))
```

---


# Transformations

```{r echo=FALSE, message=FALSE}
ggplot(df,aes(x=t,y=dlny))+
  geom_line(size=1,color="skyblue")+
  labs(y=expression(paste(Delta,"ln",y[t],sep="")))+
  theme_classic()+
  theme(axis.title = element_text(size=18),axis.text = element_text(size=14),axis.text.y = element_text(angle=90,hjust=0.5))
```
